Towards a Complete Model of the Negative Selection Process in Humans

### 2023-05-18 manuscript/manuscript.tex
draft up to theory sec

\title{Towards a Complete Model of Mutation Selection Balance in Humans}

@@

Including the total asymptotic variance in reproductive success in Equation
\eqref{eq:simple_Ne} and assuming Wright--Fisher levels of non-heritable
variation (i.e. $V = 2 + 4 Q^2 V_A$),

\begin{align}
    \label{eq:main_Ne}
    N_e = \frac{N}{Q^2 V_A + 1}
\end{align}

@@

\begin{align}
    \label{eq:polygenic_Ne}
    N_e \approx N \exp\left(-\sum_{i=1}^n V_{A,i} \frac{Q_i^2}{2}\right)
\end{align}

@@

Under any selection model, the genic fitness variance created by a new mutation
(at frequency $x=\nicefrac{1}{2N}$) is $2s^2 x(1-x) \approx \nicefrac{s^2}{N}$.
For the entire population of $2N$ chromosomes, the fresh variance from mutation
each generation in segment $i$ is $V_{m,i} \approx U_is^2$ where $U_i = 2\mu
L_i$ is the diploid mutation rate per generation within the segment. Under
mutation-selection balance assumed by BGS, an $L_i$-basepair segment has genic
variance $V_{a,i}^{BGS} \approx U_i s$ (see Supplementary Materials Equation
\ref{eq:va_bgs}) and thus $\kappa_i^{BGS} = s$. Combining this with $\kappa_i =
\nicefrac{V_{m,i}}{V_{a,i}} = s$ in Equation \eqref{eq:simple_Ne}, we have

\begin{align}
    N_e = N \exp \left( - \sum_i^n \frac{\mu L_i}{s(1 + r_i(1-s)/s)^2} \right) 
\end{align}

which is exactly the genome-wide model of background selection used in previous
studies \parencite{McVicker2009-ax,Elyashiv2016-vt,Murphy2022-sj}. Thus, the
classic background selection model is a special case of the more general theory
of Santiago and Caballero (\citeyear{Santiago2016-mu}), which they have shown
previously (\citeyear{Santiago1998-bs}).

@@

Note that since mutation is the ultimate source of all mutations that fix or
are lost, $U \ge R$. Intuitively, this equilibrium genic fitness variation is a
balance between the rate of gain of genic fitness variation due to mutation and
its loss due to fixation. Our results differ from \parencite{Santiago2016-mu},
since we find from simulations their expression models the additive
\emph{genic} fitness variation, not additive genetic fitness variation (see
Section XXX), since it does not adequately capture linkage disequilibria
dynamics. Santiago and Caballero demonstrate Equation \eqref{eq:Va} through
Fisher's Fundamental Theorem of Natural Selection, but we find an alternative
proof (\cite{Higgs1995-xc}; see Supplementary Materials Section
\ref{supp:weak-strong}).

However, closed-form expressions for the rate of the ratchet $R$ have generally
been hard to find \parencite{Haigh1978-gt,Higgs1995-xc,Gessler1995-hz}. The key
insight of \parencite{Santiago2016-mu} is that the ratchet rate under draft is
given by the probability of fixation $p_f(N_e, s)$
\parencite{Kimura1962-su,Malecot1948-zv} with the draft-rescaled effective
population size, e.g. $R = N U p_f(N_e, s)$. Given this equation for the
ratchet and Equation \eqref{eq:polygenic_Ne} for $N_e$ under draft, we have a
system of non-linear equations that can be solved numerically for $N_e$ and
$R$.

\subsection*{Simulations of the Diploid Model}

>>>>

\title{Towards a Complete Model of the Negative Selection Process in Humans}

@@

Including the total asymptotic variance in reproductive success in Equation
\eqref{eq:simple_Ne} and assuming Wright--Fisher levels of non-heritable
variation (i.e. $V = 2 + 4 Q^2 V_A$), the draft-effective population size $N_d$
is

\begin{align}
    \label{eq:main_Ne}
    N_d = \frac{N}{Q^2 V_A + 1}
\end{align}

@@

\begin{align}
    \label{eq:polygenic_Ne}
    N_d \approx N \exp\left(-\sum_{i=1}^n V_{A,i} \frac{Q_i^2}{2}\right)
\end{align}

@@

Under any selection model, the genic fitness variance created by a new mutation
(at frequency $x=\nicefrac{1}{2N}$) is $2s^2 x(1-x) \approx \nicefrac{s^2}{N}$.
For the entire population of $2N$ chromosomes, the fresh variance from mutation
each generation in segment $i$ is $V_{m,i} \approx U_is^2$ where $U_i = 2\mu
L_i$ is the diploid mutation rate per generation within the segment. Under
mutation-selection balance assumed by BGS, an $L_i$-basepair segment has genic
variance $V_{a,i}^{BGS} \approx U_i s$ (see Supplementary Materials Equation
\ref{eq:va_bgs}) and thus $\kappa_i^{BGS} = s$. Substituting $V_{a,i}^{BGS}$
and $\kappa_i^{BGS}$ in Equation \eqref{eq:simple_Ne} and simplifying, we have

\begin{align}
    N_d = N \exp \left( - \sum_i^n \frac{\mu L_i}{s(1 + r_i(1-s)/s)^2} \right) 
\end{align}

which is identical to the genome-wide model of background selection used in
previous studies \parencite{McVicker2009-ax,Elyashiv2016-vt,Murphy2022-sj}.
Thus, the classic background selection model is a special case of the more
general theory of Santiago and Caballero (\citeyear{Santiago2016-mu}), which
they have shown previously (\citeyear{Santiago1998-bs}).

@@

Note that since mutation is the ultimate source of all alleles that are fixed
by the ratchet or are lost, $U \ge R$. Intuitively, this equilibrium genic
fitness variation is a balance between the rate of gain of genic fitness
variation due to mutation and its loss due to fixation. Our results differ from
\parencite{Santiago2016-mu}, since we find from simulations their expression
models the additive \emph{genic} fitness variation, not additive genetic
fitness variation (see Section XXX), since it does not adequately capture
linkage disequilibria dynamics. Santiago and Caballero demonstrate Equation
\eqref{eq:Va} through Fisher's Fundamental Theorem of Natural Selection, but we
find an alternative proof (\cite{Higgs1995-xc}; see Supplementary Materials
Section \ref{supp:weak-strong}).

However, closed-form expressions for the rate of the ratchet $R$ have generally
been hard to find \parencite{Haigh1978-gt,Higgs1995-xc,Gessler1995-hz}. The key
insight of \parencite{Santiago2016-mu} is that the ratchet rate under draft is
given by the probability of fixation $p_f(N_d, s)$
\parencite{Kimura1962-su,Malecot1948-zv} with the draft-rescaled effective
population size, e.g. $R = N U p_f(N_d, s)$. Given this equation for the
ratchet and Equation \eqref{eq:polygenic_Ne} for $N_d$ under draft, we have a
system of two non-linear equations that can be solved numerically for $N_d$ and
$R$ for each segment,

\begin{align}
  \label{eq:}
  N_d &= N \exp \left( -V_a \frac{Q^2}{2} \right) & \text{\emph{draft-effective population equation}} \\
  R &= \frac{4N_d U s}{\exp(4 N_d s)-1}  & \text{\emph{ratchet equation}}
\end{align}

where $V_a = (U-R)s$ as in Equation \eqref{eq:Va}.

% where,

% \begin{align}
%   V_a &= (U - R(x)) s & \text{\emph{fitness variance equation}} \\
%   Q^2 &= \frac{2}{(1-Z)(2-(2-M)Z)} & \text{\emph{linkage inflation factor}} \\
%   Z &= 1 - \frac{Us}{U-{R(x)}} & \text{\emph{}} \text{\emph{variance decay rate}}
% \end{align}

\subsection*{Segment Simulations of the Diploid Model}

Given that Santiago and Caballero's (\citeyear{Santiago2016-mu}) model is a
haploid-model for a single region under equilibrium negative selection, we
first sought to confirm it could accurately predict the reduction in realistic
forward-time simulations of diploid populations with recombination. Using SLiM
\parencite{Haller2019-vu,Haller2023-uk} we simulated a region of $10^5$
basepairs under varying levels of mutation and selection. We find a close
correspondence between the observed and predicted reduction in effective
population size $B=\nicefrac{N_d}{N}$ over all selection and mutation
parameters including weak selection (Figure 1A), in contrast to classic BGS
theory. Furthermore, to investigate whether this accuracy was caused by the
model capturing the equilibrium fitness variance and ratchet rate, we also
measured these throughout the simulation. Again, we find their theory
accurately predicts both the ratchet rate (Figure 1B) and the genic fitness
variance (Figure 1C). Additionally, these figures provide further intuition for
the underlying negative selection process. When selection is strong the ratchet
rate is zero, as there is no probability that strongly deleterious can fix.
However, around $2N_e s \approx 1$, the ratchet begins clicking as $p_F > 0$.
Ultimately, the ratchet rate reaches the neutral mutation rate, as $V_a = 0$
and $N_d = N$ (e.g. no draft). Likewise, the additive genic fitness variance
matches the level predicted by BGS ($Us$), but begins to diverge when the
ratchet begins clicking.

\subsection*{Chromosome-wide Simulations and Models of Negative Selection}

Given the accuracy of the SC16 model in predicting the reduction in effective
population size and ratchet rate when equilibrium negative selection processes
were acting in single segment, we extended their model so that it could be
applied to whole-genome data.
